Abstractions and exemplars: the measure noun phrase alternation in German

### 2017-12-29 Latex/sections/2descriptive.tex
Revisions.

\section{Case in German measure NPs}
\label{sec:germanmeasurenps}

@@

@@

I now turn to some more subtle issues related to the measure noun case alternation in order to delimit the scope of the study.
First, there is a claim found in some grammars that a generic nominative, accusative, and even dative on the kind noun are used instead of the genitive (\PGCa) or case agreement (\NACa).
Overviews can be found in \cite{Hentschel1993} and \cite{Zimmer2015}.
Sentence (\ref{ex:intro:gennom}) shows a putative generic nominative on the kind noun inside an accusative MNP.

@@

\subsection{Prototype and exemplar effects}
\label{sec:analyses}

This section briefly reviews existing analyses of the \PGCa\ vs.\ \NACa\ alternation and related issues.
I also develop my own analysis and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.

.%
\footnote{A reviewer mentioned that she or he had the impression that dialectal variation is also a factor influencing the alternation.
This is definitely true as some dialects (such as Alemannic) tend to have no genitive at all.
While this is an interesting aspect for further research, the main obstacle is that there are no corpora of German which are both large enough and annotated with reliable meta data about regional variants.
In the annotation of the corpus study, documents obviously written in a regional variant were excluded.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\section{Modelling the measure noun alternation}
\label{sec:germanmeasurenps}

@@

\subsubsection{A closer look at the syntax of the alternating constructions}
\label{sec:syntax}

In order to understand the MNP alternation, it is vital to consider how in the relevant syntactic structure [N\Sub{1}~[A~N\Sub{2}]\Sub{NP\Sub{2}}]\Sub{NP\Sub{1}} (as instantiated in the \NACa\ and the \PGCa), the adjective is morpho-syntactically ambiguous between a determiner and a modifying adjective.
To show this, the strong\slash weak inflection patterns of adjectives needs to be taken into account.
In NPs with a strongly inflected determiner, attributive adjectives inflect according to the massively syncretic \textit{weak} pattern.\footnote{The weak pattern has only two suffixes.
The structural cases (nominative and accusative) in the singular are marked with \textit{-e}, the oblique cases (dative and genitive) as well as the entire plural are marked with \textit{-en}.
The only exception to this rule is the masculine accusative singular, which is also marked with \textit{-en}.}
If there is no determiner (as is the case in the alternating constructions), attributive adjectives inflect like determiners themselves.
This is called the \textit{strong} inflectional pattern.
For example, in the dative (governed here by the preposition \textit{mit}), the strong suffix \textit{-em} is on the determiner in (\ref{ex:starkschwach:a}) and the adjective bears the weak suffix \textit{-en}. 
In (\ref{ex:starkschwach:b}), the strong suffix \textit{-em} appears on the adjective because there is no determiner.%
\footnote{In the masculine and neuter singular genitive, the strong and weak forms are indistinguishable.
Since the alternation itself does not occur in the genitive (see Section~\ref{sec:descriptive}), this does not create any complications for the present study.}

\begin{exe}
  \ex\label{ex:starkschwach} 
  \begin{xlist}
    \ex{\label{ex:starkschwach:a}\gll mit ein\textbf{-em} stark\textbf{-en} Kaffee\\
                                      with a strong coffee\\}
    \ex{\label{ex:starkschwach:b}\gll mit stark\textbf{-em} Kaffee\\
                                      with strong coffee\\}
  \end{xlist}
\end{exe}

Thus, the adjectives in the \NACa\ and the \PGCa\ have properties of adjectives as well as determiners.
On the one hand, they are lexical adjectives and function as attributive modifiers.
On the other hand, they are inflected like determiners, and they are the leftmost element in the NP, which is typical of determiners.
This unusual double nature of adjectives in NPs without determiners leads to a plausible interpretation of the pattern shown in Table~\ref{tab:constructions}.
Whenever speakers classify the adjective in the kind noun phrase more as a determiner, they have to use the \PGCa\ because, if there is a determiner, the PGC is the only option.
When they classify the adjective more as an adjective, the kind NP has no determiner, and they have to use the \NACa.%
\footnote{While the generative analysis presented in \cite{Bhatt1990} cannot properly deal with probabilistic effects, Bhatt comes close to this interpretation by analysing the kind NP in the PGC as a DP and in the NAC as an NP.}
This morpho-syntactic ambiguity means that the \NACa\ is in fact a \NACb\ in disguise, and the \PGCa\ is a \PGCd\ in disguise.
This explains why the alternation arises in the first place.
In Section~\ref{sec:analyses}, I will argue that the NAC and the PGC constructions also have semantically distinguishable prototypes, and that the alternation between \PGCa\ and \NACa\ is an alternation between these prototypes.

@@

I now finally turn to some more subtle issues related to the measure noun case alternation in order to delimit the scope of the study.
First, there is a claim found in some grammars that a generic nominative, accusative, and even dative on the kind noun are used instead of the genitive (\PGCa) or case agreement (\NACa).
Overviews can be found in \cite{Hentschel1993} and \cite{Zimmer2015}.
Sentence (\ref{ex:intro:gennom}) shows a putative generic nominative on the kind noun inside an accusative MNP.

@@

\subsection{What controls the MNP alternation?}
\label{sec:analyses}

In this section, I develop my analysis of the MNP alternation and the appropriate hypotheses for the empirical studies presented in Sections~\ref{sec:corpusstudies} and \ref{sec:experimental}.
I also review some existing analyses of the alternation and related issues.
It should be remembered from Section~\ref{sec:syntax} that thanks to the syntactic ambiguity of strongly inflected adjectives, the \PGCa\ is plausibly a \PGCd\ in disguise and the \NACa\ a \NACb\ in disguise.

\subsubsection{Prototype effects}
\label{sec:prototypeeffects}

My analysis is based on the idea that the PGC prototype express a pseudo-partitive where two discernible entities -- the measure and the substance -- are referenced.
The NAC prototype, on the other hand, expresses a mere quantity where the measure is not referenced as an entitiy in its own right.
Prototypical exemplars are given in (\ref{ex:prototypes1}) for the PGC and (\ref{ex:prototypes2}) for the NAC.

\begin{exe}
  \ex\label{ex:prototypes} 
  \begin{xlist}
    \ex{\label{ex:prototypes1}\gll Sie nahmen [einen Löffel [irgendeiner Medizin]\Sub{Gen}]\Sub{Acc}.\\
    they took a spoon some medication \\
    \trans They took a spoon(ful) of some medication.
  }
    \ex{\label{ex:prototypes2}\gll Sie kauften [drei Liter [Öl]\Sub{Acc\slash caseless}]\Sub{Acc}.\\
    they bought three liters oil \\
    \trans They bought three liters of oil.
  }
  \end{xlist}
\end{exe}

While the \PGCd\ in (\ref{ex:prototypes1}) allows an interpretation where speakers conceptualise the substance itself, \ie the medication, and a spoon used to take a quantity from the medication, the \NACb\ in (\ref{ex:prototypes2}) does not allow such an interpretation.
While in the given examples, the effect is strongly supported by the choice of the measure noun lemmas, I argue that the meaning of the prototypes is independent of this.%
\footnote{A reviewer asked for a specification of the \textit{schematic meaning} of the superordinate constructions in terms of \citet{Langacker1987}.
However, none of the constructions (\PGCd, \PGCa, \NACb, and \NACa) are exclusively associated with one of the discussed meanings, and thus I do not see how one could specify schemas, for which ``membership is not a matter of degree'' and whose properties are ``fully compatible with all the members of the category'' \citep[371]{Langacker1987}.
A modelling in terms of a probabilistic similarity effects is more appropriate.
See also the argument in \citealp[70--71]{Taylor2003} and my discussion in Section~\ref{sec:cogocl}.}
The independent meanings of the prototypes are, as I will show, a result of diachronic developments and grammaticalisation processes.
Furthermore, I argue that the prototypical meanings are reflected in their usage patterns, which will be tested in the corpus study in Section~\ref{sec:corpusstudies}.

I begin by showing how the two meanings of the constructions emerge as a consequence of grammaticalisation processes of partitives and similar constructions.
It is often assumed that pseudo-partitives and quantity constructions arise as a form of grammaticalised partitives (\eg\ \citealp[536--539]{Koptjevskaja2001} for Finnish and Estonian, \citealp[559]{Koptjevskaja2001} for European languages in general).
The grammaticalisation paths uncovered by \citet[esp.\ 526--530]{Koptjevskaja2001} are relevant for the case at hand.
The grammaticalisation path can start out (in some languages) with constructions involving two referential nouns (not even necessarily forming a single and contiguous NP) and a \textit{separative} meaning as in \textit{(cut) two slices from the cake} \citep[535]{Koptjevskaja2001}.
In this type of construction, it is most obvious that two separate referents (in the given example the cake and the slices) are conceptualised.
The \textit{part-of} meaning of true partitives as in \textit{a slice of the cake} represents the first stage of a development wherein the measure noun can already lose some semantic content, when for example words like \textit{bite} are no longer necessarily interpreted as a piece literally bitten out of something.
The pseudo-partitive stage finally instantiates a \textit{quantity-of} relation, potentially even leading to fully grammaticalised quantifiers such as \textit{a lot}.
In German, the PGC is clearly the older construction \citep{Zimmer2015}.
As predicted by the grammaticalisation pattern just described, it still has the potential to form a true partitive (if the kind noun is definite).
Conversely, the NAC lacks this ability to form true partitives, and has gone further down the grammaticalisation path.
It is thus not surprising that it has lost (at least prototypically) the semantics which allows both the measure and the substance to be conceptually accessible as independent referents.
As a consequence, we should expect the NAC constructions to be typical hosting constructions for more strongly grammaticalised measure nouns.
For example, highly grammaticalised non-referential nouns like \textit{Gramm} `gram' and \textit{Meter} `metre' should occur proportionally more often in NAC constructions than in PGC constructions.
If such preferences can actually shown in usage patterns, it would lend strong support for the hypothesised difference in the meanings of the prototypes.
In the corpus study, measure lemmas will therefore be annotated with appropriate semantic class labels to check whether semantic classes of measure nouns have different affinities to the two variants.

A second preference should also be observable as a consequence of the different meanings.
As described above, the grammaticalisation path leads from NPs denoting individuated objects standing in a \textit{part-of} relation to a construction with a more diffuse \textit{quantity-of} relation.
Both types of relations can be numerically quantified -- inasmuch as a precise number of \textit{parts} or a numerically exact \textit{quantity} can be specified.
However, it is much more typical of quantities to be specified with numerical precision.
This is most obviously so with the highly grammaticalised physical measure nouns like \textit{centilitre} which are very typically used with exact numerals instead of unspecific quantifiers, although both options are available in principle (\textit{three centilitres} vs.\ \textit{several centilitres}).
Since the \NACa\ is more closely associated with the \textit{quantity-of} relation, cardinals as attributes of the measure noun are expected to have a higher proportional frequency in the \NACa.
For illustration, (\ref{ex:determiners}) shows the expected alternants under this hypothesis.

\begin{exe}
  \ex\label{ex:determiners} 
  \begin{xlist}
    \ex{\label{ex:determiners:nac}\gll [[Drei Centiliter]\Sub{Nom} [heißer Rum]\Sub{Nom}]\Sub{Nom} sind genug.\\
                                       three centilitres hot rum are enough\\
				     \glt Three centilitres of hot rum is enough.}
    \ex{\label{ex:determiners:pgc}\gll [[Einige Centiliter]\Sub{Nom} [heißen Rums]\Sub{Gen}]\Sub{Nom} sind genug.\\
                                       some   centilitres hot    rum  are   enough\\
				     \glt A few centilitres of hot rum is enough.}
  \end{xlist}
\end{exe}

In (\ref{ex:determiners:nac}), the measure noun is modified by a cardinal \textit{drei} `three', and hence the \NACa\ is preferred.
In (\ref{ex:determiners:pgc}), the measure noun is modified by a non-cardinal determiner \textit{einige} `some', and the \PGCa\ is preferred.
Especially with exact physical measure nouns (like \textit{centilitre}), exact numerical quantification is invited.
By hypothesis, however, this goes beyond a selection effect tied to measure lemmas, and cardinal quantifiers are expected to co-occur relatively more often with the \NACa.

Finally, it was found that the \PGCa\ is more typical of higher stylistic levels (distinctly edited, closer to the non-regional standard, more formal) and\slash or even exclusive to written language (see \citealp[320--323]{Hentschel1993}).
This is not surprising, given that the genitive -- an intrinsic part of the PGC -- is generally underrepresented in colloquial vernacular variants of German as a result of a diachronic process wherein many (but by no means all) uses of the genitive are being replaced by other cases or periphrastic constructions \citep{FleischerSchallert2011}.
Under an integral view of prototypes, which incorporate effects related to larger contexts, registers, and styles, such preferences can be part of what defines the construction prototypes, and the \PGCa\ should occur proportionally more often in more elaborate styles closer to the standard.%
\footnote{A reviewer mentioned that she or he had the impression that dialectal variation is also a factor influencing the alternation.
This is definitely true as some dialects (such as Alemannic) tend to have no genitive at all.
While this is an interesting aspect for further research, the main obstacle for the present study is that there are no corpora of German which are both large enough and annotated with reliable meta data about regional variants.
In the annotation of the corpus study, documents obviously written in a regional variant were excluded.}

\subsubsection{Items, exemplar effects, and hierarchical models}
\label{sec:itemandexemplareffects}

As discussed in Section~\ref{sec:cogocl}, prototype-based and exemplar-based approaches are merely the endpoints of a spectrum of theories allowing abstraction in cognitive representations to varying degrees.
While the prototypes for the PGC and the NAC were clearly specified with reference to abstract meaning in Section~\ref{sec:prototypeeffects}, at least one exemplar-driven similarity effect can also be expected to influence the MNP alternation.
The measure and kind noun lemmas which occur in the alternating constructions obviously also occur in the non-alternating constructions.
The relative frequency with which they occur in these stable cases -- where choosing an alternative is impossible -- could thus be a factor influencing the alternating, less stable case.
Should this be confirmed, it would be highly implausible to conceive of such effects as prototype effects.
In the corpus study reported in Section~\ref{sec:corpusstudies}, a measure quantifying this influence will therefore be included as the \textit{attraction strength}.

It should be noted that such and exemplar-type effect would have to exist in addition to the predicted semantic prototype effect for measure lemmas described in Section~\ref{sec:prototypeeffects}, namely the effect of semantic classes of measure nouns.

To control for the possibility that a potential lemma attraction strength effect is merely an artefact of lemma-specific tendencies, such simple lemma-specific tendencies will have to be accounted for in the modelling as well.
Additionally,

This leads to a more complex model structure if a generalised linear mixed model (GLMM) is specified, as is customary in alternation modelling (see \citealp{Gries2015} for an argument in favour of such models in corpus linguistic studies; \citealp{GelmanHill2006} (especially part 2) for a comprehensive text book on the subject).
A GLMM models the influence which several variables (\textit{predictors}) have on the probability that an alternant is chosen.
The so-called \textit{fixed effects} are assumed to be fixed population parameters which quantify the strength and the direction of the effect which, for example, a specific feature of the syntactic context or the stylistic level of the text have on the choice of the alternants.
The so-called \textit{random effects} are not fixed population parameters, but they vary by group.
The simplest random effect is a varying intercept, which predicts for groups of observations defined by lemmas, genres, speakers, etc. a constant term to be added to the model.
In a truly hierarchical model, however, the varying intercept itself comes with an additional linear model where second-level fixed effects predict the group-level effect.
To illustrate, the simple model specification in (\ref{eq:glmm}) is a hierarchical logistic regression model.

\begin{equation}
  Pr(y_i=1)=logit^{-1}(\beta^0_{j[i]}+\beta^1\cdot x^1_i+\beta^2\cdot x^2_i)
  \label{eq:glmm}
\end{equation}

The varying intercept $\beta^0_{j[i]}$ adds a constant term for group $j$ to which $i$ belongs (encoded here as $j[i]$, a notation borrowed from \citealp{GelmanHill2006}).
Each $\beta$ is a fixed-effect coefficient, and each $x$ is an observation-level variable.
(Here, two fixed effects are included for illustration purposes.)
The observations (for example, the lines of a concordance) are indexed by $i$ and the groups\slash random intercepts by $j$.
Let us say the groups are defined by lemmas and we also include lemma frequency as a control variable.
In this case, a second-level model would be given similar to (\ref{eq:level2}).

\begin{equation}
  \beta^0_j\sim\mathcal{N}(\gamma_0+\gamma^1\cdot u_j, \sigma)
  \label{eq:level2}
\end{equation}

This says that the intercepts follow a normal distribution with a standard deviation of $\sigma$ and a mean predicted from the model $\gamma_0+\gamma^1\cdot u_j$, where $\gamma^0$ is the second-level intercept, other $\gamma$s the second-level coefficients, and each $u$ a second-level predictor variable.
If $\beta^0_{j[i]}$ were a lemma random effect and we also included, for example, lemma frequency as a control variable, then $u_j$ could be the lemma frequency predictor which helps to predict the lemma frequency instead of just having simple per-lemma constants.
